On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-08-23 Manuscript/manuscript.tex
Minor change

Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
His disqualification in the final heat by reacting in 0.099 seconds, just 0.001
seconds faster than the 0.1 threshold set by the International Association of
Athletics Federations (IAAF), was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification
(\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}).
The crowd erupted with jeers
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.

@@

Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined reaction times for nine male athletes.  The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by IAAF,
\citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Unfortunately, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.
It is unknown how many athletes since the 2009 study could have potentially
benefited from this, but it is probable that there are others besides Allen who
were disqualified. Additionally,
\citet{komi2009iaaf} recommended that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

@@

The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championships to check for any abnormalities, and we approach
this in two different ways.  First we model the data, which are
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with two random effects 
one effect for the year and a second effect for the
heat within each year.
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Second, we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing reaction times from the same athlete across multiple competitions we 
are able to determine if the 2022 reaction times differed significantly from 
those of other competitions.
Our investigation into reaction times is important not only because of its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times, which is the goal of
every sprinter who steps onto the track.

The rest of this paper is organized as follows. Section~\ref{sec:Data} describes 
how data was collected and begins to detail the generalized linear mixed-effect 
(GLMM) model that is further developed in Section~\ref{sec:Methods}. The 
signed-rank test that analyzes reaction times for the same athletes across
different competitions is
also described in Section~\ref{sec:Methods}.  Section~\ref{sec:Results}
 provides the results of the two analyses and ends 
by questioning the reaction time barrier.  Lastly, Section~\ref{sec:concludingremark}
discusses the impact and limitations of the paper.

@@

There are two types of data that are explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed in
the 2022 World Championships and another competition (either 2019 World 
Championships or a 2022 national competition typically held within three months 
of the 2022 World Championships). 
\eds{is this true? are you still using the 2019 world championships data for the
rank-based analyses, or only the national competitions?}
\of{Yes the ranked based comparison has 2 subsections: 2022 national vs 2022
international and 2019 international vs 2022 international.  Should I instead say
that there are three types of data across 2 sections or something like that?}
\eds{I think this is fine here, but it is not at all clear that there are two
subsections in the rank-based results for 2022 national vs 2022
international and 2019 international vs 2022 international.}
\of{Currently under the Methods/Results section I have GLMM and rank based comparison
do you think I need subsections under rank based comparison, or is that too much
compartmentalization?  Or should I instead say there are 3 types of data that are
used in two different methods, or something like that?} \eds{I prefer
subsections within rank-based comparison.}
The first data set is used for the GLMM analyses described in 
Section~\ref{sec:glmm}, while the second collection of data is used for the 
rank-based comparisons described in Section~\ref{sec:rank}.

@@

For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together which increases the sample size which is 
advantageous because in some years there are very few finals observations, such
as in 2022 there were only five data points due to two disqualifications and one
athlete who did not compete.  We also look at data can also look at data that
excludes 2022 to see how our analysis differs based on the exclusion of one
particular year of interest.

The data under consideration is best visualized using using a boxplots, as
shown in Figure~\ref{fig:Boxplot}.  It is clear that the reaction times in the 
2022 boxplot for both 
the men's finals and men's pooled semifinal and final data are lower (indicating 
faster reaction times)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
{brosnan2017effects} found median reaction times of 0.156 seconds 
for men when looking at data from 1999 to 2014.
We can also see from Figure~\ref{fig:Boxplots} that the reaction times clearly
differ from year to year. The venue of the World Championships changes each 
year, so weather or climate related factors such as humidity, precipitation, 
elevation, may indeed play a role, but additionally, both the technology and 
false start penalization rules have changed over the study period.

\begin{figure}[tbp]
  \centering
  \includegraphics{Boxplot}
  \caption{The reaction times from 1999 to 2022 for the men's final (top) and
  the men's semifinals and finals (bottom).}
  \label{fig:Boxplot}
\end{figure}

@@

\subsubsection{2022 National Competitions}\label{sec:datanational}
From June 23-26 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World 
Championships being held at the same venue in August. Thus, we can establish a 
baseline for the four United States athletes who competed in at least one round 
of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen.   
Every athlete reacted faster in all the World Track and Field Championships 
races compared to the United States Track and Field Championships races. The 
dataset of the four United States athletes is too small to perform any 
meaningful statistical analysis, but the data was expanded to include reaction 
times for 11 athletes who competed at national competitions (many of which were 
held from May-July of 2022). This data is difficult to find as it was taken from 
various websites and was at times difficult to interpret as the data if often 
recorded in the language of the country. \of{I don't know how to say this}. Each
athlete was designated to be a cluster and the stage of competition (national or international) was specified for each competition within the cluster (athlete).
Thus, this is subunit level grouping where we assign the treatment: national or
international within each cluster.  Cluster sizes range from three to six, with
one cluster of size three, six clusters of size four, two clusters of size five
and two clusters of size six.  We chose to include preliminary rounds in this
section as well as in section~\ref{sec:data2019} in order to increase the size of our
data.  Without data from the preliminary rounds, the number of observations
dropped from 65 to 26 and the number of athletes included in the analysis
decreased from 15 to 9.

\eds{Can you show some descriptive statistics for this data?  Total number of 
clusters? Average number of international vs national competitions across 
athletes/clusters?}
\eds{The following paragraph still needs to be smoothed (the paragraph that was 
here and the paragraph that was in the intro still need to be integrated into 
one (or two) cohesive paragraph(s) that accurately describe the two types of 
data that will be used for two rank-based analyses.}
\of{Does this work?}
\eds{Yes, but you still need figure or descriptive statistics for this data.}

\subsubsection{2019 World Track and Field Championships}\label{sec:data2019}
The 2019 World Track and Field Championship data is included in the larger dataset
described in section~\ref{sec:dataworld}, but we can also use it similarly to
how we used national level in section~\ref{sec:datanational}.  We can compare
the times of athletes who participated in the 2019 and 2022 World Track and Field
Championships to see if there were lower times in 2022.  While it is theoretically
possible that athletes got faster between 2019 and 2022, it is important to
remember that these are some of the fastest sprinters on the planet, and thus
improving on elite times is not easy.  Once again we repeat a similar procedure
as in section~\ref{sec:datanational}; we assign each athlete to be a cluster and
within each cluster assign times to be from either the 2019 World Championships
or the 2022 World Championships.  In this analysis, we look at data from 15
athletes, where cluster size ranges from two to six.   There were two clusters 
of size two, three clusters of size three, five clusters of size four, two 
clusters of size five and three clusters of size six.

@@

Figure~\ref{fig:Scatterplot2019vs2022} shows the distribution of times from the
2019 and 2022 World Championships.  
\eds{Please reorganize the labels so they are either in one of the following 
orders:
2019 Heat, 2019 Semi, 2019 final, 2022 Heat, 2022 Semi, 2022 final  OR
2019 Heat, 2022 Heat, 2019 Semi, 2022 Semi, 2019 final, 2022 final.  
(I am not sure which is better, but perhaps the latter?)}
This data was only for athletes who competed
in at least one race at each championship, however it is possible for athletes
to have competed in one race in 2019 and two races in 2022.  The 2022 World
Championship Heats and Semifinals both had on average lower reaction times than
their respective 2019 races, and the 2022 finals were substantially faster. Also
worth noting is that the fastest reaction time in the 2022 Finals and 2022
Semifinals both came from Devon Allen, but the difference between 0.101 and 0.099
seconds was the difference in what caused him to be disqualified. \eds{check 
these numbers? 0.099?}\of{resolved}

@@

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one captures the venue effect in
each year and the other captures the heat effect within each year.
Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function,
$\gamma_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $\delta_{i/j}$ is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

To better understand this model, we can identify the gamma model in terms of the
commonly used shape and scale parameters. For a gamma distribution with
shape~$a$ and scale~$b$, we have mean $\mu = ab$ and variance $\nu = ab^2$. The
variance as a function of the mean~$\mu$ in the GLM sense is
$\text{var}(\mu) = \mu^2 / a$, which implies that the dispersion parameter is
$\phi = 1 / a$. Therefore, for the gamma mixed-effect model, the conditional
gamma distribution of the reaction time $Y_{ij}$ given the venue
effect~$v_i$ and heat effect~$h_{i/j}$
has shape $a = 1 / \phi$ and scale
$b = g^{-1}(\alpha + v_i + h_{i/j}) \phi$. The marginal
distribution of $Y_{ij}$ is a scale-mixture of gamma distributions, which can be
easily simulated from once the parameters are estimated. Many
random numbers generated from the fitted mixture distribution can be used to
approximate the probability of observing a reaction time faster than any given
threshold.  We are specifically interested in the probability of a reaction time
being less than 0.1 seconds in order to gauge if that is a reasonable 
disqualification barrier.

@@

With our small sample size, the asymptotic normal distribution may not be
reliable, so we use 1000 random permutations to simulate the observed test 
statistic: $W^*$ given number of times to obtain the p-value of said test 
statistics. This method is available from the \texttt{clusWilcox.test()} function
with \texttt{method = `ds'} and \texttt{exact = TRUE} from R package
\texttt{clusrank} \citep{jiang2020wilcoxon}.

@@

A summary of the two effect model is shown in Table~\ref{tab:Gamma_parameters}
for the pooled datasets and shows the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022. We choose to only look at the pooled datasets, as R displayed
warnings when creating the finals only models.  The two effect model with venue
and heat effect was the best model, as proven by an ANOVA test and decrease in
AIC.  For the entire pooled dataset; the AIC for the venue effect only model was
$-1847.9$, the heat effect only model AIC was $-1894.4$ and the both effect model,
which we will present throughout this section AIC was $-1898.9$. Thus, the two
effect model is superior and will be presented throughout the rest of the paper.
\of{How specific do I need to be in this section? Will people who read this paper
run the code to see the p value or should I say it directly and how do I say it?}

\begin{table}
  \centering
  \caption{Summary statistics for the two effect GLMM on various data sets. SE
  is short for standard error.} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & AIC & $\sigma_h^2$ & $\sigma_v^2$ & $\alpha$ & $\delta_{i/j}$ \\
      \midrule
      Old Pooled & 0.060 & 0.034 & $-1.865$ & 0.139 \\
      All Pooled & 0.059 & 0.041 & $-1.880$ & 0.014 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

Two points of this section are to determine how extreme a reaction time below
0.1 seconds is, and to determine how the exclusion of 2022 could alter results.
Using the gamma function with link set to log, the probability of the reaction 
time being less than 0.1 seconds can be calculated in R using the methods 
described in section~\ref{sec:glmm}. A simulation size of $10,000,000$ was 
chosen to try and estimate the probability of observing a reaction time below 
two thresholds: 0.1 and 0.08. When looked at the All Pooled datset, a probability
of $5.15\cdot10^{-3}$ corresponds to roughly 1 in every 195 starts
being below 0.1 seconds.  When we exclude 2022, the probability of observing a
reaction time below 0.1 seconds decreased to $4.15\cdot10^{-3}$, or 1 in every
240 starts.
\of{I feel like with the two effect model the probability change is not as
different as it was previously}

@@

We can use the same model but alter our interpretation slightly to select a 
reaction time barrier given the probability of observing a time less than the 
barrier.  For example, if we want a 0.1\% chance of observing an illegally fast 
reaction time for the men's pooled dataset that includes 2022, then the reaction
time barrier should be 0.089 seconds, as shown in Table~\ref{tab:Sim_time}.  We
can repeat this calculation for each dataset and for various probability levels,
for example a probability of 0.0001 (one disqualification for every 10,000 starts)
results in a reaction time barrier of 0.079 for the men's pooled dataset.  
Tables~\ref{tab:Sim_time} shows that the inclusion of 2022 results in a lower
reaction time barrier, further showing the claims above that the reaction times
in 2022 were low.

@@

The rank based methods described in are used in section~\ref{sec:rank_methods}
to determine if the reaction times for athletes who had competed at other 
championships differed significantly from the 2022 World Championships.  This is
a two group comparison performed two times: once comparing 2019 World
Championships to 2022 World Championships and then comparing 2022 national
championships to the 2022 World Championships.

Table~\ref{tab:Clusrankresults} shows the results generated from four tests in R.
We use a \texttt{clusWilcox.test()} with \texttt{exact = TRUE} and 
\texttt{B = 1000} indicating that the random permutation size was chosen to be 
1000.  As the sample sizes are small: 65 and 49, we use \texttt{exact = TRUE} as
that is preferred.  In both cases, the national versus international comparison
produced stronger p values than the 2019 versus 2022 comparison, with the
national vs international \texttt{exact = TRUE} test producing a p value of 
$2.2\cdot10^{-16}$, which is the smallest possible value acheivable in R.  This
is a highly significant test result and shows that athletes who competed at the
2022 Track and Field World Championships were on average, significantly faster
at reacting than they were a few months prior.  For the 2019 versus 2022
same athlete comparison with \texttt{exact = TRUE} argument, the p value was
significant at $\alpha = 0.01$, however for when \texttt{exact = FALSE}, the p
value was higher and only significant at $\alpha = 0.05$. Nonetheless, these
tests help to show that athletes competing at the 2022 Track and Field World
Championships had faster reaction times than when they had competed previously.

>>>>

Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
His disqualification in the final heat by reacting in 0.099 seconds, just 0.001
seconds faster than the 0.1 threshold set by the International Association of
Athletics Federations (IAAF), was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification
(\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}).
The crowd erupted with jeers
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.
\eds{He is not competing this year, correct? If yes, we may need to rephrase.}

@@

Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined reaction times for nine male athletes.  The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by IAAF,
\citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Unfortunately, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.
It is unknown how many athletes since the 2009 study could have potentially
benefited from this, but it is probable that there are others besides Allen who
could have. Additionally,
%were disqualified. Additionally,
\citet{komi2009iaaf} recommended that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

@@

The objective to this paper is to analyze the reaction times at the 2022 World 
Track and Field Championships to check for any abnormalities, and we approach
this in two different ways.  First we model the data, which are
the reaction times of every non-disqualification since the 1999 World 
Championships, using a gamma linear mixed-effects model with two random effects 
one effect for the year and a second effect for the
heat within each year.
Based on the fitted model, we can then estimate the probability of a reaction 
time being below a certain threshold (e.g., 0.1). 
Second, we compare reaction times for athletes who have competed
at additional competitions to the 2022 World Track and Field Championship.  By
comparing reaction times from the same athlete across multiple competitions we 
are able to determine if the 2022 reaction times differed significantly from 
those of other competitions.
Our investigation into reaction times is important not only because of its
implications for future races, but also because reaction times have
been shown to affect the overall time of the race \citep{delalija2008reaction}.
Thus, faster reaction times lead to faster overall times, which is the goal of
every sprinter who steps onto the track.
\eds{Since the World Championships are happening again now, is data from this 
year already available? If it is available, we would need to justify its omission
in the current analysis.
If it is not available, but becomes available while under review, would we 
reanalyze the data?  Does it seem likely our conclusions would change based on
what is going on this year?}

The rest of this paper is organized as follows. Section~\ref{sec:Data} describes 
how data was collected and begins to detail the generalized linear mixed-effect 
(GLMM) model that is further developed in Section~\ref{sec:Methods}. The 
signed-rank test that analyzes reaction times for the same athletes across
different competitions is
also described in Section~\ref{sec:Methods}.  Section~\ref{sec:Results}
 provides the results of the two analyses and ends 
by questioning the reaction time barrier.  Lastly, Section~\ref{sec:concludingremarks}
discusses the impact and limitations of the paper.

@@

There are two types of data that are explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed in
the 2022 World Championships and another competition (either 2019 World 
Championships or a 2022 national competition typically held within three months 
of the 2022 World Championships). 
%\eds{is this true? are you still using the 2019 world championships data for the
%rank-based analyses, or only the national competitions?}
%\of{Yes the ranked based comparison has 2 subsections: 2022 national vs 2022
%international and 2019 international vs 2022 international.  Should I instead say
%that there are three types of data across 2 sections or something like that?}
%\eds{I think this is fine here, but it is not at all clear that there are two
%subsections in the rank-based results for 2022 national vs 2022
%international and 2019 international vs 2022 international.}
%\of{Currently under the Methods/Results section I have GLMM and rank based comparison
%do you think I need subsections under rank based comparison, or is that too much
%compartmentalization?  Or should I instead say there are 3 types of data that are
%used in two different methods, or something like that?} \eds{I prefer
%subsections within rank-based comparison.}
The first data set is used for the GLMM analyses described in 
Section~\ref{sec:glmm}, while the second collection of data is used for the 
rank-based comparisons described in Section~\ref{sec:rank}.

@@

For analysis purposes, we will often pool reaction times of men's final and 
men's semifinal heats together which increases the sample size. This is 
advantageous because in some years there are very few finals observations, such
as in 2022, when there were only five data points due to two disqualifications 
and one
athlete who did not compete.  We also consider data that
excludes 2022 to see how our analysis differs based on the exclusion of one
particular year of interest.

The data under consideration is summarized using a boxplots in 
Figure~\ref{fig:Boxplot}.  It is clear that the reaction times in the 
2022 boxplot for both 
the men's finals and men's pooled semifinal and final data are lower (indicating 
faster reaction times)
than in previous years.  The median for the men's pooled data in 2022 was 0.129
and the median for the men's final data in 2022 was 0.126.  For comparison,
\cite{brosnan2017effects} found median reaction times of 0.156 seconds 
for men when looking at data from 1999 to 2014.
We can also see from Figure~\ref{fig:Boxplot} that the reaction times clearly
differ from year to year. The venue of the World Championships changes each 
year, so weather or climate related factors such as humidity, precipitation, 
elevation, may indeed play a role, but additionally, both the technology and 
false start penalization rules have changed over the study period.

\begin{figure}[tbp]
  \centering
  \includegraphics{Boxplot}
  \caption{The reaction times from 1999 to 2022 for the men's final (top) and
  the men's semifinals and finals (bottom). 
	\eds{there is only one plot here, not two. If you are keeping only one, the
	text and caption need to be modified accordingly}}
  \label{fig:Boxplot}
\end{figure}

@@

\subsubsection{2022 National Competitions}\label{sec:datanational}
From June 23-26, 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World 
Championships being held at the same venue in August. Thus, we can establish a 
baseline for the four United States athletes who competed in at least one round 
of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen.   
Every athlete reacted faster in all the World Track and Field Championships 
races compared to the United States Track and Field Championships races. The 
dataset of the four United States athletes is too small to perform any 
meaningful statistical analysis, but the data was expanded to include reaction 
times for 11 more athletes who competed at other national competitions (many of which were 
held from May-July of 2022). This data was difficult to find, as the data is not
centrally located but instead found on different national websites (often 
recorded in the national language without translation, and hence not used). 
\of{I don't know how to say this}. \eds{is my change correct?} Each athlete was 
designated as a ``cluster'' and the stage of competition (national or international) 
was specified for each competition within the cluster (athlete).
Thus, this is subunit level grouping where we assign the treatment, national or
international, within each cluster.  Cluster sizes range from three to six, with
one cluster of size three, six clusters of size four, two clusters of size five,
and two clusters of size six.  We chose to include preliminary rounds in this
section as well as in Section~\ref{sec:data2019} in order to increase the size 
of our data.  Without data from the preliminary rounds, the number of observations
dropped from 65 to 26 and the number of athletes included in the analysis
decreased from 15 to 9.

%\eds{Can you show some descriptive statistics for this data?  Total number of 
%clusters? Average number of international vs national competitions across 
%athletes/clusters?}
%\eds{The following paragraph still needs to be smoothed (the paragraph that was 
%here and the paragraph that was in the intro still need to be integrated into 
%one (or two) cohesive paragraph(s) that accurately describe the two types of 
%data that will be used for two rank-based analyses.}
%\of{Does this work?}
%\eds{Yes, but you still need figure or descriptive statistics for this data.}

\subsubsection{2019 World Track and Field Championships}\label{sec:data2019}
The 2019 World Track and Field Championship data is included in the larger dataset
described in Section~\ref{sec:dataworld}, but we can also use it similarly to
how we used national level in Section~\ref{sec:datanational}.  We can compare
the times of athletes who participated in the 2019 and 2022 World Track and Field
Championships to see if there were lower times in 2022.  While it is theoretically
possible that athletes got faster between 2019 and 2022, it is important to
remember that these are some of the fastest sprinters on the planet, and thus
improving on elite times is not easy.  Once again we repeat a similar procedure
as in Section~\ref{sec:datanational}; we assign each athlete to be a cluster and
within each cluster label times to be from either the 2019 World Championships
or the 2022 World Championships.  In this analysis, we look at data from 15
athletes, where cluster size ranges from two to six.   There were two clusters 
of size two, three clusters of size three, five clusters of size four, two 
clusters of size five, and three clusters of size six.

@@

Figure~\ref{fig:Scatterplot2019vs2022} shows the distribution of times from the
2019 and 2022 World Championships.  
\eds{Please reorganize the labels so they are either in one of the following 
orders:
2019 Heat, 2019 Semi, 2019 final, 2022 Heat, 2022 Semi, 2022 final  OR
2019 Heat, 2022 Heat, 2019 Semi, 2022 Semi, 2019 final, 2022 final.  
(I am not sure which is better, but perhaps the latter?)}
This data was only for athletes who competed
in at least one race at each championship, however it is possible for athletes
to have competed in one race in 2019 and two races in 2022.  The 2022 World
Championship Heats and Semifinals both had on average lower reaction times than
their respective 2019 races, and the 2022 finals were substantially faster. Also
worth noting is that the fastest reaction time in the 2022 Finals and 2022
Semifinals both came from Devon Allen, but the difference between 0.101 and 0.099
seconds was the difference in what caused him to be disqualified. 
%\eds{check these numbers? 0.099?}\of{resolved}

@@

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the final model we chose was a gamma mixed-effect 
model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one captures the venue effect in
each year and the other captures the heat effect within each year.
Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. 
\eds{I was talking to my husband about this. Could the heat effect be justified 
to capture the variability in the amount of time between the ``set'' and the 
gun going off (assuming this is different from heat to heat)? E.g., quicker 
gunshots after the ``set'' could lead to slower reaction times (or the reverse), 
but this effect would be shared across all athletes in the same heat.}
In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function,
$\gamma_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $\delta_{i/j}$ is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

To better understand this model, we can identify the gamma model in terms of the
commonly used shape and scale parameters. For a gamma distribution with
shape~$a$ and scale~$b$, we have mean $\mu = ab$ and variance $\nu = ab^2$. The
variance as a function of the mean~$\mu$ in the GLM sense is
$\text{var}(\mu) = \mu^2 / a$, which implies that the dispersion parameter is
$\phi = 1 / a$. Therefore, for the gamma mixed-effect model, the conditional
gamma distribution of the reaction time $Y_{ijk}$ given the venue
effect~$v_i$ and heat effect~$h_{i/j}$
has shape $a = 1 / \phi$ and scale
$b = g^{-1}(\alpha + v_i + h_{i/j}) \phi$. The marginal
distribution of $Y_{ijk}$ is a scale-mixture of gamma distributions, which can be
easily simulated from once the parameters are estimated. Many
random numbers generated from the fitted mixture distribution can be used to
approximate the probability of observing a reaction time faster than any given
threshold.  We are specifically interested in the probability of a reaction time
being less than 0.1 seconds in order to gauge if that is a reasonable 
disqualification barrier.

@@

With our small sample size, the asymptotic normal distribution may not be
reliable, so we use 1000 random permutations to simulate the observed test 
statistic: $W^*$ given number of times to obtain the p-value of said test 
statistics. 
\eds{I do not understand what is meant in the text after the colon in the 
previous sentence.  Please rephrase.}
This method is available from the \texttt{clusWilcox.test()} function
with \texttt{method = `ds'} and \texttt{exact = TRUE} from R package
\texttt{clusrank} \citep{jiang2020wilcoxon}.

@@

A summary of the two effect model is shown in Table~\ref{tab:Gamma_parameters}
for the pooled datasets and shows the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022. We choose to only look at the pooled datasets, as R displayed
warnings when creating the finals only models.  The two effect model with venue
and heat effect was the best model, as indicated by an ANOVA test 
\eds{I think it may be better to call this a likelihood ratio test rather than
an ANOVA test, assuming that is the formal test being used} and decrease in
AIC.  For the entire pooled dataset, the AIC for the venue effect only model was
$-1847.9$, the AIC for the heat effect only model $-1894.4$, and the AIC for the 
two effect model, which we will present throughout this section, was $-1898.9$. 
Thus, the two
effect model is superior and will be presented throughout the rest of the paper.
\of{How specific do I need to be in this section? Will people who read this paper
run the code to see the p value or should I say it directly and how do I say it?}
\eds{If you are doing a likelihood ratio test, I suggest including the p-value 
and the degrees of freedom associated with the test.}

\begin{table}
  \centering
  \caption{Summary statistics for the two effect GLMM on various data sets. SE
  is short for standard error. \eds{data in table needs to be revisited and table
	needs to be referenced within the text. Also, it needs to be clarified 
	somewhere that ``Old Pooled'' excludes 2022 and ``All Pooled'' includes 2022.}} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & AIC & $\sigma_h^2$ & $\sigma_v^2$ & $\alpha$ & $\delta_{i/j}$ \\
      \midrule
      Old Pooled & 0.060 & 0.034 & $-1.865$ & 0.139 \\
      All Pooled & 0.059 & 0.041 & $-1.880$ & 0.014 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

Two aims of this section are to determine how extreme a reaction time below
0.1 seconds is, and to determine how the exclusion of 2022 could alter the results.
Using the gamma function with link set to log, the probability of the reaction 
time being less than 0.1 seconds can be calculated in R using the methods 
described in Section~\ref{sec:glmm}. A simulation size of $10,000,000$ was 
chosen to estimate the probability of observing a reaction time below 
two thresholds: 0.1 and 0.08. When examining the All Pooled dataset, a probability
of $5.15\cdot10^{-3}$ corresponds to roughly 1 in every 195 starts
being below 0.1 seconds.  When we exclude 2022, the probability of observing a
reaction time below 0.1 seconds decreased to $4.15\cdot10^{-3}$, or 1 in every
240 starts.
\of{I feel like with the two effect model the probability change is not as
different as it was previously}

@@

We can use the same model but alter our interpretation slightly to select a 
reaction time barrier given the probability of observing a time less than the 
barrier.  For example, if we want a 0.1\% chance of observing an illegally fast 
reaction time for the men's pooled dataset that includes 2022, then the reaction
time barrier should be 0.089 seconds, as shown in Table~\ref{tab:Sim_time}.  We
can repeat this calculation for each dataset and for various probability levels,
for example a probability of 0.0001 (one disqualification for every 10,000 starts)
results in a reaction time barrier of 0.079 for the men's pooled dataset.  
Table~\ref{tab:Sim_time} shows that the inclusion of 2022 results in a lower
reaction time barrier, further showing the claims above that the reaction times
in 2022 were low.

@@

The rank based methods described in Section~\ref{sec:rank}
to determine if the reaction times for athletes who had competed at other 
championships differed significantly from the 2022 World Championships.  This is
a two group comparison performed two times: once comparing 2019 World
Championships to 2022 World Championships and then comparing 2022 national
championships to the 2022 World Championships.
